---
title: "Task1"
author: "Marcos Crespo"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
    keep_tex: yes
fontsize: 10pt
editor_options: 
  chunk_output_type: console
link-citations: yes
---

**PARA PONER LA BIBLIOGRAFIA**
**PONER EL INDICE EN SIDEBAR**
**VER MAS ESTILOS DE HIGHLIGHT**
bibliography: rep.bib
biblio-style: apalike
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("qmnist_nist.RData")
```
```{r libraries, echo=FALSE, warning=FALSE}
# dplyr
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
library(dplyr)

# glmnet
if (!requireNamespace("glmnet", quietly = TRUE)) {
  install.packages("glmnet")
}
library(glmnet)
```

# Introduction



# Data insights
The MNIST database is a widely used dataset in the field of machine learning and computer vision. It stands for Modified National Institute of Standards and Technology database. It consists of a large collection of handwritten digits. Each observation consists of the label (the intended writen number), the writter id and the $28*28$ greyscale grid values for the image itself. Possible representation of a $0$ could be:


```{r visualize, echo=FALSE}
show_digit <- function(x, col = gray(255:1 / 255), ...) {
l <- sqrt(length(x))
image(matrix(as.numeric(x), nrow = l)[, l:1], col = col, ...)
}

show_digit(train_nist[1,3])
```

This version of the dataset includes approximately 30,000 training images and 30,000 testing ones. These images are normalized and centered, making them a standard benchmark for evaluating the performance of some algorithms in tasks such as digit recognition and classification[^1].

[^1]: For some application and references see **VER COMO CITAR** my final degree thesis (VER SI PONER TFG) **PONER URL** .

In the class assignment, we are asked to implement a ridge logistic model for classifying the digits 4 and 9, so our next step is to filter only the data with labels 4 and 9. 
```{r labels}
# Filter our original data
train <- train_nist %>%
  filter(digit %in% c(4, 9))

train$digit <- droplevels(train$digit)
```
After running this filtering we end up with 6011 observations of numbers 4 and 9

```{r}
summary(train[,c(1,2)])
```

# The model

Ridge regression is a method of estimating the coefficients of multiple-regression models in scenarios where the independent variables are highly correlated.[1] It has been used in many fields including econometrics, chemistry, and engineering.*[WIKIPEDIA]*

*This may be the current situation for the correlation thing (breve explicaion del modelo?)*

The ridge model can be found in `glmnet` package as a subcase for $\alpha=0$. We will be trying to fit ridge a model, with a cross-validated-chosen λ penalty 

```{r ridge2}
# Lambda is a tuning parameter that can be chosen by cross-validation, using as
# error the MSE (other possible error can be considered for generalized models
# using the argument type.measure)

# 10-fold cross-validation. Change the seed for a different result
set.seed(12345)
kcvRidge <- cv.glmnet(x = as.matrix(train[,3]), y = train[,1], alpha = 0, standardize = FALSE, family = "binomial", nfolds = 3)

# The lambda that minimizes the CV error is
kcvRidge$lambda.min
## [1]26.46999

# Equivalent to
indMin <- which.min(kcvRidge$cvm)
kcvRidge$lambda[indMin]
## [1] 26.46999

# The minimum CV error
kcvRidge$cvm[indMin]
## [1] 0.1163279
min(kcvRidge$cvm)
## [1] 0.1163279

# Potential problem! Minimum occurs at one extreme of the lambda grid in which
# CV is done. The grid was automatically selected, but can be manually inputted
range(kcvRidge$lambda)
## [1]     25.52821 255282.09651

#not in the margins so we dont need to redefine the grid for lambda


# But the CV curve is random, since it depends on the sample. Its variability
# can be estimated by considering the CV curves of each fold. An alternative
# approach to select lambda is to choose the largest within one standard
# deviation of the minimum error, in order to favor simplicity of the model
# around the optimal lambda value. This is known as the "one standard error rule"
kcvRidge$lambda.1se
## [1] 2964.928

# Location of both optimal lambdas in the CV loss function in dashed vertical
# lines, and lowest CV error and lowest CV error + one standard error
plot(kcvRidge)
abline(h = kcvRidge2$cvm[indMin] + c(0, kcvRidge$cvsd[indMin]))




```
```{r bet plot}
plot(coef(kcvRidge, s = "lambda.1se")[-1])

#Intercept exlcuded because of scale. Comment this
```

```{r labels2}
# Filter our original data
test <- test_nist %>%
  filter(digit %in% c(4, 9))

test$digit <- droplevels(test$digit)
```

```{r predict}
plot(kcvRidge$glmnet.fit, label = TRUE, xvar = "lambda")
abline(v = log(c(kcvRidge$lambda.min, kcvRidge$lambda.1se)))



# The model associated to lambda.1se (or any other lambda not included in the
# original path solution -- obtained by an interpolation) can be retrieved with
predict(kcvRidge, type = "coefficients", s = kcvRidge$lambda.1se)


# Predictions
predictions <- predict(kcvRidge, type = "response", s = "lambda.min",
        newx = as.matrix(test$px))


# Convert predicted probabilities to class labels (4 or 9)
predicted_classes <- ifelse(predictions > 0.5, 9, 4)

# Evaluate the predictions
accuracy <- mean(predicted_classes == test$digit)

# Print the accuracy
cat("Accuracy on test data:", accuracy)

```





